Project Protocol

### 2019-10-21 protocol.Rmd
sample size calculations

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

By chance, the data plot will be detected with probability, $1/12=0.083$. This corresponds to about 2 detections out of 20, generally this would be $n/12$ detections. Let's suppose the alternative hypothesis has probability of detection $p_A (>1/12)$. If we expect the data to be detected twice as often as by chance, $p_A=1/6$. To determine the probability of observing $n/12$ detections, given $p_A$, we can use a binomial distribution, where $k=ceiling(n/12)$, to calculate the power:

$$P(X>k|p=p_A,n) = {n \choose k} p_A^k (1-p_A)^{n-k}$$

If $p_A=1/6$ we would need almost 100 evaluations for each lineup to expect at least $n/12$ detections. With $p_A=0.25$ approximately 50 evaluations should yield sufficient detections. We will use a pilot study to estimate $p_A$.

@@

```{r echo=FALSE, eval=FALSE}
# Code to compute the power for given sample size and alternative p
library(tidyverse)
library(nullabor)
library(viridis)
n <- 20:100
k <- ceiling(n/12)
pA <- seq(1/12, 1/2, 0.01)
g <- expand.grid(n, k, pA)
pow <- tibble(n=g[,1], k=g[,2], pA=g[,3]) 
pow <- pow %>% 
  mutate(prob = pbinom(k, n, pA, lower.tail=FALSE)) 
ggplot(pow, aes(x=n, y=pA, colour=prob, group=pA)) + 
  geom_point() +
  scale_colour_viridis_c()
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  error = FALSE, 
  message = FALSE)
```

@@

By chance, the data plot will be detected with probability, $1/12=0.083$. This corresponds to about 2 detections out of 20 evaluatoins, and generally this is $n/12$ detections from $n$ evaluations. Let's suppose the alternative hypothesis has probability of detection $p_A (>1/12)$. If we expect the data to be detected twice as often as by chance, $p_A=1/6$. To determine the probability of observing $n/12$ detections, for one lineup, given $p_A$, we can use a binomial distribution, where $k=ceiling(n/12)$, to calculate the power:

$$P(X\geq k|p=p_A,n) = \sum_{j=k}^{n}{n \choose j} p_A^j (1-p_A)^{n-j}$$

@@

```{r echo=FALSE, eval=TRUE, fig.height=5, fig.width=6, out.width="70%"}
# Code to compute the power for given sample size and alternative p
library(tidyverse)
library(nullabor)
library(viridis)
library(knitr)
n <- 20:50
pA <- seq(1/12, 1/3, 0.002)
g <- expand.grid(n, pA)
pow <- tibble(n=g[,1], k=as.integer(ceiling(g[,1]/12)), pA=g[,2]) 
pow <- pow %>% 
  mutate(prob = pbinom(k-1, n, pA, lower.tail=FALSE)) 
ggplot(pow, aes(x=n, y=pA, fill=prob, group=pA)) + 
  geom_tile() +
  scale_fill_viridis_c("power") +
  geom_hline(yintercept=0.1666667, colour="black") +
  ylab("detect rate (pA)") + xlab("sample size (n)") +
  theme_bw()
pow %>% filter(near(pA, 0.167, 0.001)) %>% filter(prob>0.9) %>% top_n(n=10) %>% kable(digits=3)
```

The table above shows the sample sizes yielding above 90% chance of detection ($p_d$) in a single lineup assuming $p_A=1/6=0.167$ we would need approximately 50 evaluations for each lineup.

If 4 lineups are evaluated for each trend model, the chance that none produce a detection is $(1-p_d)^4$, which for $p_D>0.9$ is effectively 0.
